Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-23 src/tex/ms.tex
[showyourwork] overleaf sync

The key assumption underpinning our methodology — that individual stars are independently sampled from their respective stellar environments — is foundational for tractability. However, this assumption warrants closer examination. In reality, stellar abundances are correlated due to shared star formation histories, cumulative enrichment, and dynamical interactions. Future efforts should explore methodologies capable of incorporating such correlations to further refine the accuracy of SBI, potentially through hierarchical modeling or graph-based methods.

@@

Our results underscore the transformative advantages of SBI over traditional methods like Hamiltonian Monte Carlo (HMC), marking a paradigm shift in galactic parameter inference:
\begin{itemize}
    \item Orders-of-magnitude speed-up: SBI dramatically reduces computational requirements, achieving runtime improvements exceeding 1,000-fold compared to HMC. For instance, while HMC requires $\sim 40\,\mathrm{hours}$ to infer parameters from just 200 stars, SBI completes inference on thousands of stars in mere minutes. This efficiency makes SBI uniquely suited for analyzing the massive datasets expected from next-generation spectroscopic surveys.
    \item Scalability: SBI’s amortized nature allows it to scale seamlessly with the size of the dataset. By training a neural posterior estimator once, the method can be applied repeatedly at virtually no additional computational cost. This scalability is essential for leveraging the millions of stars that future surveys like 4MOST and WEAVE will provide, enabling precise population-level inferences.
    \item Robustness to model misspecifications: Unlike HMC, which shows significant biases when faced with discrepancies between model assumptions and data, SBI demonstrates remarkable robustness. \textcolor{blue}{Even under conditions of mismatched yield tables (should be updated)} or data generated from hydrodynamical simulations, SBI provides accurate and reliable results. This robustness ensures SBI’s applicability in real-world scenarios where exact model fidelity cannot be guaranteed.
\end{itemize}

>>>>

The key assumption underpinning our methodology -- that individual stars are independently sampled from their respective stellar environments -- is foundational for tractability. However, this assumption warrants closer examination. In reality, stellar abundances are correlated due to shared star formation histories, cumulative enrichment, and dynamical interactions. Future efforts should explore methodologies capable of incorporating such correlations to further refine the accuracy of SBI, potentially through hierarchical modeling or graph-based methods.

@@

Our results underscore the transformative advantages of SBI over traditional methods like Hamiltonian Monte Carlo (HMC), marking a paradigm shift in galactic parameter inference:
\begin{itemize}
    \item Orders-of-magnitude speed-up: SBI dramatically reduces computational requirements, achieving runtime improvements exceeding 1,000-fold compared to HMC. For instance, while HMC requires $\sim 40\,\mathrm{hours}$ to infer parameters from just 200 stars, SBI completes inference on thousands of stars in mere minutes. This efficiency makes SBI uniquely suited for analyzing the massive datasets expected from next-generation spectroscopic surveys.
    \item Scalability: SBI's amortized nature allows it to scale seamlessly with the size of the dataset. By training a neural posterior estimator once, the method can be applied repeatedly at virtually no additional computational cost. This scalability is essential for leveraging the millions of stars that future surveys like 4MOST \citep{4most} and WEAVE \citep{weave} will provide, enabling precise population-level inferences.
    \item Robustness to model misspecifications: Unlike HMC, which shows significant biases when faced with discrepancies between model assumptions and data, SBI demonstrates remarkable robustness. \textcolor{blue}{Even under conditions of mismatched yield tables (should be updated)} or data generated from hydrodynamical simulations, SBI provides accurate and reliable results. This robustness ensures SBI's applicability in real-world scenarios where exact model fidelity cannot be guaranteed.
\end{itemize}
